Patchwork: Incremental Computation for Abstract Interpretation

### 2019-02-05 paper.tex
Added some citations to section 3

Both Flix and IncA require that all functions over lattices are \textit{monotone}. This is a non-trivial restriction in the context of static analysis, as many standard analyses use non-monotonic widening functions to ensure termination. All of these are inadmissible for Flix or IncA. Adapton has no monotonicity restriction.\\

There is an important sense in which the approach taken by IncA lacks 'history' in the sense of Adapton. We illustrate via an example. Consider an arbitrary lattice with elements $a, b, c, d$ such that the following holds: $a$ and $b$ are not ordered, $a \sqcup c = d$, $b \sqcup c = d$, $c \neq d$. Assuming that $a$ and $c$ are in the initial set of facts, an analysis run will add $d$ to the database. Now assume we make the following edit: delete $a$ and insert $b$. This is not a monotonically increasing replacement, so IncA will transitively delete all facts depending on $a$, including $d$ and any facts derived from $d$. Now after adding $b$, in the incremental run, the analysis will quickly re-discover $d$. However, since $d$ was already deleted from the database, all consequents of $d$ must be completely re-computed. Because IncA cannot 'look ahead' to see that $d$ will be re-discovered from an inserted fact, work that could have been saved is lost. This is a specific consequence of the general problem that deleting all facts deriving from a deleted fact is an over-approximation that is necessary for the Datalog approach to be sound. In contrast, Adapton memoizes the calls which use $d$ in further derivations. When $a$ is deleted $d$ is stale, but as soon as $d$ is re-discovered, the any downstream computations memo-match and do not need to be re-computed.

>>>>

Both Flix and IncA require that all functions over lattices are \textit{monotone}. This is a non-trivial restriction in the context of static analysis, as many standard analyses use non-monotonic widening functions to ensure termination. All of these are inadmissible for Flix or IncA. In the presence of non-monotone functions, IncA could not guarantee that monotonically increasing replacements would not require deletions. Therefore, to achieve from-scratch consistency, they would have to fallback to the naive incremental strategy. Adapton has no monotonicity restriction.\\

There is an important sense in which the approach taken by IncA lacks 'history' in the sense of Adapton. We illustrate via an example. Consider an arbitrary lattice with elements $a, b, c, d$ such that the following holds: $a$ and $b$ are not ordered, $a \sqcup c = d$, $b \sqcup c = d$, $c \neq d$. Assuming that $a$ and $c$ are in the initial set of facts, an analysis run will add $d$ to the database. Now assume we make the following edit: delete $a$ and insert $b$. This is not a monotonically increasing replacement, so IncA will transitively delete all facts depending on $a$, including $d$ and any facts derived from $d$. Now after adding $b$, in the incremental run, the analysis will quickly re-discover $d$. However, since $d$ was already deleted from the database, all consequents of $d$ must be completely re-computed. Because IncA cannot 'look ahead' to see that $d$ will be re-discovered from an inserted fact, work that could have been saved is lost. This is a specific consequence of the general problem that deleting all facts deriving from a deleted fact is an over-approximation that is necessary for the Datalog approach to be sound. In contrast, Adapton memoizes the calls which use $d$ in further derivations. When $a$ is deleted $d$ is stale, but as soon as $d$ is re-discovered, the any downstream computations memo-match and do not need to be re-computed.\\

 There have been some attempts to alleviate the over-deletion problem (B/F alg \url{https://www.cs.ox.ac.uk/boris.motik/pubs/mnph15incremental-BF.pdf}) (hybrid alg \url{https://arxiv.org/pdf/1711.03987.pdf}). This appears to be an improvement to Delete/Rederive algorithms, but not in all cases, and Adapton should still give you this behavior effectively for free.\\
